Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-31 manuscript/draft.Rmd
some manuscript changes

Developing forecasts for biodiversity is important for conservation, management,
and evaluating how well current models characterize natural systems. While there
are an increasing number of studies forecasting biodiversity, there is little
information available on how well these forecasts work because these studies
fail to follow good practice for making and evaluating forecasts. Most diversity
forecasts do not conduct meaningful evaluations of the forecasts, fail to
account for uncertainty, and do not use time-series data that captures the
actual dynamics being studied. Using hindcasting we made and evaluated `r
numbers$n_forecasts` forecasts for breeding bird species richness at `r
numbers$N_predict_sites` sites throughout the continental United States,
following best practices for forecasting. Forecasts were made for up to a decade
after the final training observations. Our results show that while all of the
models we evaluated produce reasonable forecasts, none of them consistently
outperforms the simple baseline of long-term average richness. This means that
at present, our best forecast for future richness is simply to assume that it is
unchanging. This calls into question the efficacy of current modeling
approaches. Most of the best practices implemented in this study directly
influence the outcome of the forecasts and result in different conclusions from
those made in previous studies. This emphasizes the importance of following good
practice in making ecological forecasts and of starting to actively make more
forecasts in such a way that they can be quickly evaluated and improved on.

@@

Forecasting the future state of ecological systems is increasingly important for
planning, management, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001; @houlahan2017;
@dietze2017]. One area of particular interest for forecasting is biodiversity
since it is important for ecosystem function, central to conservation planning,
and expected to be influenced by anthropogenic global change [@cardinale2012;
@diaz2015; @tilman2017]. Anticipating potential changes in biodiversity is
crucial for large scale management and conservation, and for addressing debates
regarding whether local scale diversity is declining and may continue to decline
in response to anthropogenic influences [@diaz2015; @dornelas2014].

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models.
In general, these models describe individual species' occurrence patterns as
functions of the environment [@thomas2004; @urban2015]. By plugging forecasts of
future environmental conditions into these functions, ecologists can make
predictions about where each species will occur in the future. These
species-level predictions can then be added up ("stacked") to generate predicted
richness values [e.g. @calabrese2014]. Alternatively, models that directly
relate spatial patterns of species richness to the environment have been
developed and generally perform equivalently to species distribution modeling
based methods [@algar2009; @distler2015]. This approach is sometimes referred to
as "macroecological" modeling, because it models the larger-scale pattern
(richness) more directly [@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity, little is known about how effectively we can anticipate their
dynamics. This is due in part to the long time scales over which many ecological
forecasts are applied and the resulting difficulty in assessing whether the
predicted changes occurred [@dietze2016]. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009; @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation
(Box 1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation [@clark2001,
@dietze2017]; and 3) conducting meaningful evaluations of the forecasts by
hindcasting, archiving short-term forecasts, and comparing forecasts to
baselines to determine whether the forecasts are more accurate than simply
assuming the system is basically static [@perretti2013].

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. This 30 year
time-scale has only recently become possible for large-scale time-series based
forecasting and allows us to model and assess changes a decade or more into the
future in the presence of shifts in environmental conditions on par with
predicted climate change. We compare traditional distribution modeling based
approaches to spatial models of species richness, time-series methods, and
simple baselines. All forecasts model uncertainty and observation, are evaluated
across different time lags using hindcasting, and are publicly archived to allow
future assessment. We discuss the implications of these practices for our
understanding of, and confidence in, the resulting forecasts, and how we can
continue to build on these approaches improve ecological forecasting in the
future.

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan] and its R interface [@rstan]. Because observer and
site are strongly related (observers tend to repeatedly sample the same site),
site was also included as a random effect to ensure that inferred deviations
were actually observer-related (as opposed to being related to the sites that a
given observer happened to see). The resulting model partitions the variance in
observed richness values into site-level variance, observer-level variance, and
residual variance (e.g. variation within a site from year to year). The
site-level estimates can also be used directly as the "average" baseline model
(see below). The estimated observer effects can be subtracted from the richness
values for a particular observer to provide an estimate of how many species
would have been found by a "typical" observer. To incorporate uncertainty in
these "corrected" richness values into the forecasting models we collected 500
Monte Carlo samples from the model's posterior distribution, and fit each of the
downstream models with each of the Monte Carlo samples.

@@

When comparing forecasts for richness across sites all methods performed well
(Figure 2; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites resulting in a notable clustering of predicted values near ~60 species and
the poorest performance ($R^2$ `r numbers$env_R2s`, versus XX-XX for the
within-site methods).

@@

While the models generally did well in absolute terms (Figure 2), none tended to
outperform the "average" baseline from the generalized linear mixed model
(Figure 3A). [[Stats go here]].  The auto-ARIMA was generally the
best-performing non-baseline model, but in many cases (67% of the time), the
best-fitting ARIMA model chosen by the model selection procedure had no
autoregressive, drift, or moving average terms, making it similar to the
"Average" model.

@@

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 3B). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution also caused it to have notably higher deviance (Figure
3C) than the next-worst model, even though its point estimates were not
unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. Because they avoided this assumption,
the GBM-based "macroecological" model and especially the mistnet JSDM had
reasonably well-calibrated uncertainty estimates (Figure 3B); as a result, their
relative performance was higher in terms of deviance than in terms of RMSE. The
"naive" model was the only model whose confidence intervals were too wide
(Figure 3B), which can be attributed to the rapid rate at which these intervals
expand as the time lag increases (Figure 1).

In general, each model's error increased more-or-less linearly with time,
although the slopes and initial error rates varied across models (Figure 4). The
mistnet JSDM and GBM-based macroecological model's deviance tended to rise less
rapidly over time than the baseline models' deviance did; for the ten-year
forecasts, these environment-based models outperformed the Naive model and
nearly matched the performance of the Average model.

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 5). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The Naive model was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The Average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The mistnet
JSDM was notable here because the coverage of its prediction intervals retained
very good coverage in the presence of this observer-based noise, which we
attribute to its ability to model this variation with its latent variables.

>>>>

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While there are
an increasing number of studies forecasting biodiversity, there is little
information available on how well these forecasts work. Most diversity forecasts
do not conduct meaningful evaluations of the forecasts, fail to account for
uncertainty, and do not use time-series data that captures the actual dynamics
being studied. Using hindcasting we made and evaluated `r numbers$n_forecasts`
forecasts for breeding bird species richness at `r numbers$N_predict_sites`
sites throughout the continental United States, following ten suggested best
practices for forecasting. Forecasts were made for up to a decade after the
final training observations. While all of the models we evaluated produce
reasonable forecasts, none of them consistently outperforms the simple baseline
of constant long-term average richness. At present, our best forecast for future
richness is thus to assume that it is unchanging, and current modeling
approaches do not reliably improve on this simple baseline. Most of the best
practices implemented in this study directly influence the outcome of the
forecasts and result in different conclusions from those made in previous
studies. This emphasizes the importance of following good practice in making
ecological forecasts and of starting to actively make more forecasts in such a
way that they can be quickly evaluated and improved on.

@@

Forecasting the future state of ecological systems is increasingly important for
planning, managing, and evaluating how well ecological models capture the key
ecological processes governing a system [@clark2001; @houlahan2017;
@dietze2017]. Forecasting is biodiversity is especially important, due to its
central role in conservation planning and its sensitivity to anthropogenic
effects [@cardinale2012; @diaz2015; @tilman2017]. Anticipating potential changes
in biodiversity is crucial for large scale management and conservation, and for
addressing debates regarding whether local scale diversity is declining and may
continue to decline in response to anthropogenic influences [@diaz2015;
@dornelas2014].

Previous efforts to predict future patterns of species richness, and diversity
more generally, have focused primarily on building species distributions models
(SDMs). In general, these models describe individual species' occurrence
patterns as functions of the environment [@thomas2004; @urban2015]. By plugging
forecasts of future environmental conditions into these functions, ecologists
can make predictions about where each species will occur in the future. These
species-level predictions can then be added up ("stacked") to generate predicted
richness values [e.g. @calabrese2014]. Alternatively, models that directly
relate spatial patterns of species richness to the environment have been
developed and generally perform equivalently to stacked SDMs [@algar2009;
@distler2015]. This approach is sometimes referred to as "macroecological"
modeling, because it models the larger-scale pattern (richness) directly
[@distler2015].

Despite the emerging interest in forecasting species richness and other aspects
of biodiversity, little is known about how effectively we can anticipate their
dynamics. This is due in part to the long time scales over which many ecological
forecasts are applied [and the resulting difficulty in assessing whether the
predicted changes occurred @dietze2016]. What we do know comes from a small
number of hindcasting studies, where models are built using data on species
occurrence and richness from the past and evaluated on their ability to predict
contemporary patterns [e.g., @algar2009; @distler2015]. These studies are a
valuable first step, but lack several components that are important for
developing good forecasting models and understanding how accurately these models
can predict the future. These components of good forecasting and evaluation
(Box 1) broadly involve: 1) expanding the use of data to include biological and
environmental time-series [@tredennick2016]; 2) accounting for uncertainty in
observations, processes, model choice, and forecast evaluation [@clark2001,
@dietze2017]; and 3) conducting meaningful evaluations of the forecasts by
hindcasting, archiving short-term forecasts, and comparing forecasts to
baselines to determine whether the forecasts are more accurate than simply
assuming the system is basically static [@perretti2013].

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. This 30 year
time-scale has only recently become possible for large-scale time-series based
forecasting and allows us to model and assess changes a decade or more into the
future in the presence of shifts in environmental conditions on par with
predicted climate change. We compare traditional distribution modeling based
approaches to spatial models of species richness, time-series methods, and
simple baselines (Figure 1). All forecasts model uncertainty and observation
error, are evaluated across different time lags using hindcasting, and are
publicly archived to allow future assessment. We discuss the implications of
these practices for our understanding of, and confidence in, the resulting
forecasts, and how we can continue to build on these approaches improve
ecological forecasting in the future.

@@

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. We estimated the observer
effects (and associated uncertainty about those effects) with a linear mixed
model with observer as a random effect built in the Stan probabilistic
programming language [@stan]. Because observer and site are strongly related
(observers tend to repeatedly sample the same site), site was also included as a
random effect to ensure that inferred deviations were actually observer-related
(as opposed to being related to the sites that a given observer happened to
see). The resulting model partitions the variance in observed richness values
into site-level variance, observer-level variance, and residual variance (e.g.
variation within a site from year to year; Figure 3). The site-level estimates
can also be used directly as the "average" baseline model (see below). The
estimated observer effects can be subtracted from the richness values for a
particular observer to provide an estimate of how many species would have been
found by a "typical" observer. To incorporate uncertainty in these "corrected"
richness values into the forecasting models we collected 500 Monte Carlo samples
from the model's posterior distribution, and fit each of the downstream models
with each of the Monte Carlo samples.

![](../figures/observer_predictions.png)

@@

When comparing forecasts for richness across sites all methods performed well
(Figure 4; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites resulting in a notable clustering of predicted values near ~60 species and
the poorest performance ($R^2$ `r numbers$env_R2s`, versus XX-XX for the
within-site methods).

@@

While the models generally did well in absolute terms (Figure 4), none tended to
outperform the "average" baseline from the generalized linear mixed model
(Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
best-performing non-baseline model, but in many cases (67% of the time), the
best-fitting ARIMA model chosen by the model selection procedure had no
autoregressive, drift, or moving average terms, making it similar to the
"Average" model.

@@

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 4B). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution also caused it to have notably higher deviance (Figure
4C) than the next-worst model, even though its point estimates were not
unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. Because they avoided this assumption,
the GBM-based "macroecological" model and especially the mistnet JSDM had
reasonably well-calibrated uncertainty estimates (Figure 4B); as a result, their
relative performance was higher in terms of deviance than in terms of RMSE. The
"naive" model was the only model whose confidence intervals were too wide
(Figure 4B), which can be attributed to the rapid rate at which these intervals
expand as the time lag increases (Figure 1).

In general, each model's error increased more-or-less linearly with time,
although the slopes and initial error rates varied across models (Figure 5). The
mistnet JSDM and GBM-based macroecological model's deviance tended to rise less
rapidly over time than the baseline models' deviance did; for the ten-year
forecasts, these environment-based models outperformed the Naive model and
nearly matched the performance of the Average model.

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 6). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The Naive model was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The Average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The mistnet
JSDM was notable here because the coverage of its prediction intervals retained
very good coverage in the presence of this observer-based noise, which we
attribute to its ability to model this variation with its latent variables.
